---
title: "Getting started with HoneyBADGER"
author: "Jean Fan"
date: '`r Sys.Date()`'
output: html_document
md_document:
  variant: markdown_github
vignette: |
  %\VignetteIndexEntry{Vignette Title} %\VignetteEngine{knitr::rmarkdown} \usepackage[utf8]{inputenc}
---


```{r, include = FALSE}
library(knitr)
opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = 'figure/',
    cache.path = 'cache/',
    cache = TRUE,
    out.width='800px',
    out.height='400px',
    dpi=100
)
```


```{r}

library(HoneyBADGER)

```

Load gene expression data for cells from patient MGH31 along with an expression reference.

```{r}
data(gexp)
data(ref)

require(biomaRt) ## for gene coordinates
mart.obj <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = 'hsapiens_gene_ensembl', host = "jul2015.archive.ensembl.org")
```

Make new HoneyBADGER object and initialzie the gene expression matrices. The data has already been filtered for highly expressed shared genes and scaled for library size differences so we can override the default filtering and scaling.

```{r, fig.width=12, fig.height=4}
hb <- new('HoneyBADGER', name='MGH31')
hb$setGexpMats(gexp, ref, mart.obj, filter=FALSE, scale=FALSE)
hb$plotGexpProfile(setOrder=TRUE)
```

Now we can model the gene expression variance and use a HMM to identify regions affected by CNVs.

```{r}
hb$setMvFit()
hb$calcGexpCnvBoundaries(init=TRUE)

## double check what CNVs were identified
bgf <- get('bound.genes.final', slot(hb, '.xData'))
genes <- get('genes', slot(hb, '.xData'))
regions.genes <- range(genes[unlist(bgf)])
print(regions.genes)
```

Retest all identified CNVs on all cells, cluster cells on the final posterior probability, and plot.

```{r}
hb$retestIdentifiedCnvs()

## look at final results
retest <- get('results', slot(hb, '.xData'))[['gene-based']]
results.amp <- do.call(rbind, lapply(retest, function(x) x[[1]])) 
results.del <- do.call(rbind, lapply(retest, function(x) x[[2]])) 

## filter to regions with at least some highly confident cells
vi <- rowSums(results.amp > 0.8) > 10 | rowSums(results.del > 0.8) > 10
print(regions[vi])

results <- rbind(results.amp[vi,], results.del[vi,])
hc <- hclust(dist(t(results)), method='ward.D')
```


```{r, fig.width=12, fig.height=4}
## plot all chromosomes
hb$plotGexpProfile(cellOrder=hc$labels[hc$order])
## plot identified CNV region only
hb$plotGexpProfile(cellOrder=hc$labels[hc$order], region=regions[vi])
```

```{r}
## summarize results
hb$summarizeResults(geneBased=TRUE, alleleBased=FALSE)
```

```{r, fig.width=12, fig.height=4}
## visualize as heatmap 
hb$visualizeResults(geneBased=TRUE, alleleBased=FALSE, hc=hc)
```

We can also analyze using allele information. Load the allele data and add them to our HoneyBADGER object.

```{r}
data(r)
data(cov.sc)

hb <- new('HoneyBADGER', name='MGH31')
hb$setAlleleMats(r.init=r, n.sc.init=cov.sc, n.cores=1, het.deviance.threshold=0.1)

library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
hb$setGeneFactors(txdb)
```


```{r, fig.width=12, fig.height=4}

hb$plotSmoothedAlleleProfile(cellOrder=hc$labels[hc$order])
hb$plotAlleleProfile(cellOrder=hc$labels[hc$order])

```

Now run our allele HMM to identify putative regions affected by deletions or LOH.

```{r}
hb$calcAlleleCnvBoundaries(init=TRUE)

## double check what CNVs were identified
bsf <- get('bound.snps.final', slot(hb, '.xData'))
snps <- get('snps', slot(hb, '.xData'))
regions.snp <- range(snps[unlist(bsf)])
print(regions.snp)
```

We can again retest.

```{r}
hb$retestIdentifiedCnvs(retestBoundGenes=FALSE, retestBoundSnps=TRUE)

retest <- get('results', slot(hb, '.xData'))[['allele-based']]
results <- do.call(rbind, lapply(retest, function(x) x))


```{r, fig.width=12, fig.height=4}
hb$plotAlleleProfile(cellOrder=hc$labels[hc$order])

hb$plotAlleleProfile(cellOrder=hc$labels[hc$order], region=regions.snp[1])
hb$plotAlleleProfile(cellOrder=hc$labels[hc$order], region=regions.snp[5])
```

```{r}
hb$summarizeResults(geneBased=FALSE, alleleBased=TRUE)
```

```{r, fig.width=12, fig.height=4}
## visualize as heatmap 
hb$visualizeResults(geneBased=FALSE, alleleBased=TRUE, hc=hc)
```
